\hspace{1cm} Do syntactic and semantic similarity lead to interference effects? \newline Evidence from self-paced reading and event-related potentials using German

### 2023-12-25 paper/main.tex
minor edit to abstract and removed comments

\abstract{Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed to interpret the sentence. If these retrieval cues match with not only the features on the noun that is the retrieval target but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
% We are explaining what we did, not summarizing every detail in the paper:
%Previously, behavioral studies have consistently found syntactic interference at the verb, but semantic interference was only found in later regions or not at all. 
% What we did:
We present large-sample self-paced reading (N = 774) and event-related potentials experiments (N = 103) using a well-established design \citep{vandyke07,mertzen}, to investigate interference due to syntactic and semantic cues in German. 
% Not needed for an abstract, plus sounds boring:
%The aim of our experiments was to gain precise estimates of the effects of syntactic and semantic interference to inform theory building.
% What we found:
A Bayes factor analysis showed that 
both the self-paced reading times as well as event-related potentials provided clear evidence for a semantic interference effect. In the self-paced reading experiment, interference due to the semantic cues caused a reading time slow-down starting at the distractor. 
% The reader does not even know what that is:
%The best explanation for this effect is encoding interference. 
The event-related potentials showed a more negative N400 amplitude at the critical verb, i.e., the retrieval site, for high vs.\ low semantic interference.
% Irrelevant in an abstract:
 %and no such effects at preceding words. 
 % No new information in this sentence, a waste:
 %Therefore, the event-related potentials were modulated by retrieval interference.
% Said this more compactly already:
%Bayes factors for both reading times and event-related potentials provided evidence for the semantic interference effect. 
Surprisingly, in both experiments, Bayes factor analyses showed evidence against interference due to syntactic cues. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assume that both syntactic and semantic cues play an equal role in retrieval. We show through computational modeling that cue-based retrieval also shows no syntactic interference if the parser is assumed to keep track of which clause the subject occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the predictions of the cue-based retrieval model would be consistent with the observed patterns in our data.

In sum, this large-sample study of German using self-paced reading and event-related potentials suggests that although both syntactic and semantic cues drive retrieval at the verb, syntactic cues may include hierarchical syntactic information, which can render the parser immune to syntactic interference effects.

% This cannot be the main point of this paper! 
%The present findings, which are in line with various sentence processing accounts, highlight the importance of semantic cues during sentence processing.

% Highlights must be in a separate file (not in the paper itself)
\clearpage

\section{Highlights} 
% 3-5 highlights
% only 85 characters including spaces per highlight
\begin{itemize}
\item Self-paced reading and event-related potentials experiments on interference
\item Largest studies on interference during subject-verb integration (N=774, N=103)
\item Bayes factors show evidence for semantic, but against syntactic interference
\item Modeling shows that tracking of syntactic structure attenuates syntactic interference
% Wasting highlights on low information content
%\item SPR: Encoding interference lead to longer reading times starting at a distractor
%\item ERP: Time course of effect clearly indicated an effect of retrieval interference
\end{itemize}

>>>>

\abstract{Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed to interpret the sentence. If these retrieval cues match with not only the features on the noun that is the retrieval target but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
We present large-sample self-paced reading (N = 774) and event-related potentials experiments (N = 103) using a well-established design \citep{vandyke07,mertzen}, to investigate interference due to syntactic and semantic cues in German. 
A Bayes factor analysis showed that 
both the self-paced reading times as well as event-related potentials provided clear evidence for a semantic interference effect. In the self-paced reading experiment, interference due to the semantic cues caused a reading time slow-down starting at the distractor. 
The event-related potentials showed a more negative N400 amplitude at the critical verb, i.e., the retrieval site, for high vs.\ low semantic interference.
Surprisingly, in both experiments, Bayes factor analyses showed evidence against interference due to syntactic cues. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assume that both syntactic and semantic cues play an equal role in retrieval. We show through computational modeling that cue-based retrieval also shows no syntactic interference if the parser is assumed to keep track of which clause the subject occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the predictions of the cue-based retrieval model would be consistent with the observed patterns in our data.

In sum, this large-sample study of German using self-paced reading and event-related potentials suggests that, in our data, although both syntactic and semantic cues drive retrieval at the verb, syntactic cues may include hierarchical syntactic information, which can render the parser immune to syntactic interference effects.

\clearpage

\section{Highlights} 
% 3-5 highlights
% only 85 characters including spaces per highlight
\begin{itemize}
\item Self-paced reading and event-related potentials experiments on interference
\item Largest studies on interference during subject-verb integration (N=774, N=103)
\item Bayes factors show evidence for semantic, but against syntactic interference
\item Modeling shows that tracking of syntactic structure attenuates syntactic interference
\end{itemize}
